\documentclass[11pt,a4paper]{article}
\usepackage{linguex}
\usepackage{tipa}
\usepackage{apacite,pslatex}
\usepackage{titling}
\title {Nominal Derivaton in Afaan Oromoo}
\author {Tolemariam Fufa}
\date{September 2022}
\begin{document}
\newcommand\keywords[1]{%
	\begingroup
	\let\and\\
	\par
	\noindent\textbf{Keywords:}\\#1\par
	\endgroup
}
\maketitle
\begin{abstract}
	
	Afaan Oromoo Nominal derivation is not thoroughly treated in the literature. In fact, \cite{temesgen1985} mentioned some nominal derivations such as result nominals. Yet, he doesn’t determine range of variations of nominal derivations in terms of type, form and meaning. This paper is aimed to fill such research gap. The paper is divided into three parts. The first part is dedicated to identify describe agent nominals in terms of definition, type, form and meaning in Afaan Oromoo. This section discusses agent nominal in relation to semantic roles because semantic roles relate verbs with arguments; this is important as agent nominals are derived from dynamic verbs. It also describes the three types of agent nominals: The doer of an act, agents with a functional status and behavioral agent by citing Afaan Oromoo examples.The second section is devoted to event nominals. This section defines event nominals interms of argument structure and derivatioal patterns. It also considers form and meaning of event nominals of Afaan Oromoo. The third section discusses result nominals. The section defines result nominals. The section discusses whether or not result nominals differ from event nominals. The section also describes that result nominals are referential in the sense that they do not relevant to semantic roles \cite{melloni2012event}. Moreover, the section discusses forms of result nominals, specifically from which root or base result nominals derived. The section also discusses meaning ranges of result nominals.\\
		
	
\end{abstract}
\keywords{nominal, derivation, agent, event, result}
\newpage
\section{Introduction}
In Afaan Oromoo nominals can be derived from different word classes. Many abstract nouns are derived from nouns by affixing morphemes such as -ummaa \cite{gaddisa2001,temesgen1985,tolemariam2011}. For instance the following  abstract nominals are derived in such manner. \\

\ex.
\ag.
nam-ummaa \\
man-N\\
'humanity'\\

\ex.
\ag.
garb-ummaa \\
slave-N\\
'slavery'\\

\ex.
\ag.
dubart-ummaa \\
female-N\\
'feminity'\\

As shown in these examples abstract nominals namummaa 'humanity' garbummaa 'slavery' and dubartummaa 'feminity' are derived from nouns nama 'man', garba 'slave' and dubartii 'female' respectively. 


In a similar way, nouns can be derived from adjectival bases in Afaan Oromoo \cite{gaddisa2001,temesgen1985,tolemariam2011,tolemariam2009}. \\

\ex.
\ag.
gaar-ummaa \\
good-N\\
'goodness'\\

\ex.
\ag.
diim-ina \\
red-N\\
'redness'\\

\ex.
\ag.
ham-ina \\
cruel-N\\
'cruelty'\\

As shown above abstract nominals gaarummaa 'goodness', diimina 'redness' and hammina 'cruelty' are all derived from adjectival bases. 

In Afaan Orommoo deverbial nominals are also abundant. Nominals can be derived from intransitive and transitive roots. More interestingly, in this language, different nominal categories can be derived from derived verbs such as causative, middle and passive. The main focus of this paper is nominal categories which are derived from derived verbs such as causative, middle and passive. Particularly the paper focuses on four deverbial nominal derivations namely agent, event, result and patient. 

The paper is amied to answer the following research questions:\\
\begin{enumerate}
	\item What criterion do we use to identify agent, event, result and patient nominals? 
	\item From which verb stems are agent, event, result and patient nominals derived?
	\item Which nominal affixes are employed to derive agent, event, result and patient nominals?
\end{enumerate}

The nature of this research is qualitative. Data is gathered from different Afaan Oromoo books and MA theses. I used Linguistic Latex \footnote{LaTeX creates papers, presentations and artiles widely used in linguistics and other scientific researches.} free software tool for write up. 

\section{Agent Nominal}

In linguistic literature agent nominals are nominals derived from agentive verbs; they describe performers of an action \cite{huyghe2020s}. In fact, the concept agent is difficult to define. Major properties  such as animacy, control, volition, instigation, accountability, motion, etc have been long outlined and debated in the literature. Animacy, volition, and instigation has never been a global criterion for agent. Some linguists claim that agentive refers to an animate entity which happen to be the willful source or agent of the activity described by the verb \cite{cruse1973some} while others argue against this concept and on the contrary they claim that intention is not a necessary criterion of agentivity for we often do things accidentally \cite{schlesinger1989instruments,caspar2015relationship}. For my discussion I follow \cite[p-188]{huyghe2020s} general definition of agents as effectors that are prototypically, but not necessarily, animate and intentional in the sense that the participant can be volitional or original instigator that brings about something about; it refers to an action performed by an entity which is considered as employing its own energy in carrying out the action. Hence, expriencers are not considered  agents since agents are assumed to be involved in dynamic situations. 


Agent nominals are described in terms of  semantic roles. Semantic roles explain the relationship between verbs and arguments. In this sense nouns can be categorized as agent nominals if they link their referent as the agent of intrinsically specified action \cite[p-189]{huyghe2020s}. 

Two conditions must be met for typology of agent nominal. One, the presence of dynamic action in the semantic structure; and two, the description of the the agent. In this case agent nominals are assumed to be deverbal nouns that denote the agentive argument of the base verb \cite[p-189-190]{huyghe2020s}.

The range of variation of prototypical agent nominals can be classified into 3 types. The doer of an act (agent nominals that denote agents involved in a specific event), the agent of an action ( ANs that denote agents with a functional status, can be used as bare predicates) and behavioral agent (that denote agents with a propensity to do certain things or to act in a certain way, repetition of actions) \cite[190-192]{huyghe2020s}. For example, ‘savior’  is doer of an act; ‘rescue worker’ is an agent of a function; while ‘fighter’, ‘hard worker’ , etc are considered as behavioral agents.

In many languages agent nominals are identified by agent morpheme that is attached to the base fo the nominal under consideration. For example in English morphemes sucha as -er (teacher), -ist (populist), an (veteran), eer (engineer) indicate agent nominals \cite[p-186]{huyghe2020s}.

In Afaan Oromoo, as far as I know, there is no linguistic research dedicated to the description of agent nominals. Agent nominals in this language are identified by the moropheme -aa and -tuu. The morpheme -aa marks masculine whereas the morpheme -tuu marks feminine or dimunitive. In fact in Afaan Oromoo agent nominal derivation involves complex morphological processes. They are morophologically complex in the sense that they require causative or middle bases. \\

Doer of an act (derived from transitive roots/bases)\\

Agent nominals are underlyingly causers. Causer nominals are agent nominals. In Afaan Oromoo, agent nominals are created from corresponding verbs in a complex way in the sense that they involve morphological processes. Specifically, they require the causative affix to be attached (not in all cases) to which the agent nominal affix is get attached to. 

\ex.
\ag.
bar-e \\
know-3MSS\\
'He knew.'\\

\ex.
\ag.
bar-siis-aa/tuu \\
know-CAUS-NM/F\\
'teacher'\\

\ex.
\ag.
konkol:aat-e \\
roll:MID-3MSS\\
'It rolled.'\\

\ex.
\ag.
konkol:aat-is-aa/tuu \\
rool:MID-CAUS-NM/F\\
'driver'\\

\ex.
\ag.
gudd-aa \\
big-ADJ\\
'big'\\

\ex.
\ag.
gudd-is-aa/tuu \\
big-CAUS-NM/F\\
'babysitter'\\


As shown in (7) bar- ‘to know’ is the root word to which the causative morpheme – siis – is attached in (8). agent nominal morpheme – aa – or – tuu – is suffixed to the causative base. 

There are also cases where more complex morphological processes involved. For example in the following instances three derivation suffixes, namely \emph{-oom-, -s- and -aa/tuu} create agent nominal.\\

\ex.
\ag.
fir-oom-e \\
relative-MID-3MSS\\
'He became relative.'\\

\ex.
\ag.
fir-oom-s-e \\
relative-MID-CAUS-3MSS\\
'He made somebody to become relative.'\\

\ex.
\ag.
fir-oom-s-aa/tuu \\
relative-MID-NM/F\\
'socializing person'\\


Behavioral agent (derived from intransitive roots)\\

\ex.
\ag.
hat-e \\
steal-3MSS\\
'He stole (something)'\\

\ex.
\ag.
hat-tuu \\
steal-NF\\
'thief'\\

\ex.
\ag.
sob-e \\
lie-3MSS\\
'He lied.'\\

\ex.
\ag.
sob-tuu/duu \\
lie-NF\\
'liar'\\


\section{Event Nominal}
Event nominals are also known as action or process nominal in the literature \cite{siloni1997event}. Syntactically, Event nominals share thematic relations of the base verb; they inherit the argument structure of the base verb \cite{uth2015event,siloni1997event}. Semantically, they transpose the meaning of base verbs. Event nominals are “abstract nouns that ‘give a name’ to the situations (i.e.event or states) expressed by their corresponding predicates.”\cite{uth2015event}.

In Afaan Oromoo, event nominals are derived by mophemes  -aatii, -itii, \\

\ex.
\ag.
isaan karaa cuf-an \\
they.NOM road.ABS block-3PS.PER\\
'They blocked a road.'\\

\ex.
\ag.
karaa cuf-aatii isaan-ii \\
road block-N they-GEN\\
'Their road blockage'\\


As shown in (20) cuf-an ‘blocked’ is a transitive verb, karaa ‘road’ is object and isaan ‘they’ is subject. (21) is a nominal phrase derived from (20). In (21), the head noun is cuf-aatii ‘blockage’. This nominal is an event nominal that is derived from base verb cuf- ‘block’. The nominalizing affix -aatii is attached to the base verb cuf- ‘block’ to derive the event nominal cuf-aatii. The phrase ‘karaa cufaatii isaanii’ gives name to the event described by verb cuf- ‘block’ in (20). \\

\ex.
\ag.
isaan kaleessa muka mur-an \\
they.NOM yesterday tree cut-3PS.PER\\
'They cut a tree yesterday.'\\

\ex.
\ag.
muka mur-aatii kaleessaa \\
tree cut-N yesterday\\
'yesterday's tree cuttng'\\

\ex.
\ag.
nuti tulluu yaab-ne \\
we mountain climb-1PS\\
'We climbed a tree.'\\

\ex.
\ag.
tulluu yaabb-itii \\
mountain climb-N\\
'mountain climbing'\\


Event nominal derivation is not straight forward always. In some cases event nominal overlaps with result nominal as shown in the following example:\\

\ex.
\ag.
inni biiraa \textipa{\!d}ug-e \\
he.NOM beer drink-3MSS\\
'He drank beer.'\\

\ex.
\ag.
biiraa \textipa{\!d}ug-aatii \\
beer drink-N\\
'beer drinking'\\

\ex.
\ag.
(5c)\\
\textipa{\!d}ug-aatii\\
(a) event nominal \\
(b) referential (it means any kind or drink)\\

\section{Result Nominal}
As compared to agent and event nominals, result nominals are said to be non-argumental nominals. Result nominals are not explained in terms of semantic roles and verb-argument relationships. They are not formulated in the spirit of GB theory (Uth, 2015:4). Therefore, result nominals are treated from semantic point of view. That is to say the meanings of result nominals are not transposed to or from argument structure of the verb from which they have been derived. Uth claims that result nominals are purely  referential \cite[p-4]{uth2015event}. But my data shows that this is not the case. I came across result nominals which are referential as well as result nominas that are explined in terms of semantic roles. 

In Afaan Oromoo, many result nominals employ unique affixes such as -umsa and -ina \cite{gaddisa2001,temesgen1985,tolemariam2011}. The suffix -umsa is attached to root verb to derive result nominal as shown in (6c). This type of result nominals are purely referential. \\

\ex.
\ag.
k'or-umsa\\
exmine-N\\
‘examination’\\
*k'or-at-umsa\\
*k'or-siis-umsa\\

\ex.
\ag.
jeek'-umsa\\
distrub-N\\
‘distrubance’\\
*jeek'-sis-umsa\\

\ex.
\ag.
bar-umsa RESULT\\
know-N\\
‘education’\\
*bar-siis-umsa\\
*bar-at-umsa\\

\ex.
\ag.
beek-umsa\\
Know-N\\
‘Knowledge’ Result\\
*beek-sis-umsa\\ 


The morpheme -ina is attached to verbs orginated from adjectival roots. Such result nominals are said to be measure result nouns \cite{uth2015event}. The morpheme -ina can also be attached to intransitive and transitive roots to derive rusult nominals \cite{gaddisa2001,temesgen1985,tolemariam2011}. \\

\ex.
\ag.
gudd-at-e\\
big-MID-3MSS\\
'He became big'\\

\ex.
\ag.
gudd-ina\\
big-N\\
'growth'\\

\ex.
\ag.
\textipa{\!d}eer-ina\\
long-N\\
'length'\\

\ex.
\ag.
bal\textipa\textbarglotstop-at-e\\
wide-MID-3MSS\\
'width'\\

\ex.
\ag.
bal\textipa\textbarglotstop-ina\\
'wide-N\\
'width'\\

The morpheme can be attached intransitive and transitive verb roots to derive result nominals \cite{gaddisa2001,temesgen1985,tolemariam2011}.\\

\ex.
\ag.
fiig-e \\
run - 3MSs\\
'He run'\\

\ex.
\ag.
fiig-insa (fiig-ica)\\
run -N\\
'run'\\

\ex.
\ag.
diig-insa\\
destroy-N\\
'destruction'\\

\ex.
\ag.
k'ot-insa (k'otisa)\\
farm-N\\
'farm'\\

\section{Patient Nominal}
Patient is the person or entity that is affected by the action described by a verb. The causer and the controller of the action is known as agent as we have discussed in section one. The following criterion can be considered for patient: undergoes change of state, incremental theme, causally affected by another participant and stationary relative to movement of another participant \cite{barker1993nominal}. 

In Afaan Oromoo patient nominals are often derived from middle or passive bases as opposed to agent nominals which are often drived from causative stems \cite{tolemariam2009}. In this language person names clearly show the comparison of agent and patient nominal derivations. We can dare to say for every agent nominal there is one patient nominal as far as proper nouns or person names concerned. \\

\ex.
\ag.
gudd-is-aa/tuu  AGENT\\
big-CAUS-NM/F\\
'Guddisaa/One who makes it grow'\\

\ex.
\ag.
Gudd-at-aa/tuu PATIENT\\
big-MID-NM/F\\
'Guddataa/One who become big'\\

\ex.
\ag.
Gabb-is-aa/ee AGENT\\
fat-CAUS-NM/F\\
'Gabbisa/One who makes it fat'\\

\ex.
\ag.
Gabb-at-aa/tuu PATIENT\\
fat-MID-NM/F\\
'Gabbisaa/One who becomes fat'\\


\ex.
\ag.
\textipa{\!d}eer-eess-aa AGENT\\
long/tall-CAUS-N\\
'\textipa{\!d}eeressaa'/One that makes it long/tall'\\

\ex.
\ag.
\textipa{\!d}eer-at-aa PATIENT\\
long/tall-CAUS-N\\
'\textipa{\!d}eerataa'/One who becomes long/tall'

Professoion wise we can observe agent-patient contrast nominal derivations.

\ex.
\ag.
bar-siis-aa/tuu AGENT\\
know-CAUS-NM/F\\
‘Teacher’\\

\ex.
\ag.
bar-at-aa/tuu PATIENT \\
know-MID-NM/F\\
‘Student’\\

\ex.
\ag.
hojj-at-iis-aa/hojj-aciis-aa AGENT \\
work-MID-CAUS-N\\
‘Employer/Supervisor’\\

\ex.
\ag.
hojj-at-aa PATIENT\\
work-MID-N\\
‘worker’\\

Patient can also be derived from a passive stem.\\

\ex.
\ag.
gor-s-aa/tuu AGENT\\
advise-CAUS-NM/F\\
‘Advisor’\\

\ex.
\ag.
gor-s-am-aa/tuu PATIENT\\
advise-CAUS-PASS-NM/F\\
‘Advisee’\\ 



\section{Conclusion}

	Assumption 1: Agent, Event, Result and Patient nominals are deverbials.\\
Assumptioin 2: Derivational affixes such as passive, causative, transitivizer, middle create stems to which nominal affixes are attached\\
Assumption 3: Complex derivational suffixes sucha as CAUS + MID, MID + CAUS, MID + PASS, etc play roles in deriving agent , event adn patient nominals in Afaan Oromoo.
Assumption 4: Result nominals by virtue of their inherent referential character avoid verb derivation affixes such as CAUS, MID and PASS.
Assumption 5: Result nominal derivation require non-argument deverbial affixes. 
Assumption 6: I believe that analysis of derived nominal in terms of these affixes explains patterns of deverbal nominals. \\

\newpage
\bibliographystyle{apacite}
\bibliography{agent}

\newpage
\section*{Acronyms and symbols}
ABS=Absolute form\\
ADJ=Adjectivizer
c=palatal affricate\\
CAUS=Causative\\
F=Feminine\\
GEN=Genitive
M=Masculine\\
MID=Middle\\
N=Nominal\\
NF=Nominal Feminine\\
NM=Nominal Masculine\\
NOM=Nominative\\
1PS=First Plural Subject\\
3MSS=Third Masculine Singular Subject\\
3PS=Third Plural Subject \\
PER=Perfect
POSS=Possessive\\
PASS=Passive\\
\end{document}